% sections/evaluation.tex

\section{Evaluation}
\label{sec:evaluation}

We evaluate the routing system across three dimensions: signal extraction efficiency, LoRA multi-task scaling, and end-to-end routing correctness.

\subsection{Signal Extraction Latency}

\Cref{tab:signal_latency} reports median and p99 latencies for each signal type on an NVIDIA A100 GPU with ModernBERT base model.

\begin{table}[h]
\centering
\caption{Signal extraction latency by type}
\label{tab:signal_latency}
\begin{tabular}{lrrl}
\toprule
\textbf{Signal Type} & \textbf{Median} & \textbf{p99} & \textbf{Requires ML} \\
\midrule
Keyword      & $< 0.1$\,ms & $< 0.5$\,ms  & No \\
Context      & $< 0.1$\,ms & $< 0.5$\,ms  & No \\
Language     & $< 0.5$\,ms & $< 1$\,ms    & No \\
Authorization & $< 0.1$\,ms & $< 0.5$\,ms & No \\
\midrule
Embedding    & $15$\,ms    & $45$\,ms     & Yes \\
Domain       & $60$\,ms    & $120$\,ms    & Yes \\
Fact-check   & $55$\,ms    & $110$\,ms    & Yes \\
Modality     & $50$\,ms    & $100$\,ms    & Yes \\
Feedback     & $55$\,ms    & $115$\,ms    & Yes \\
Complexity   & $50$\,ms    & $105$\,ms    & Yes \\
Preference   & $55$\,ms    & $110$\,ms    & Yes \\
\bottomrule
\end{tabular}
\end{table}

Heuristic signals complete in sub-millisecond time, while ML signals range from 15--120\,ms.
With parallel evaluation, the wall-clock time is dominated by the slowest active signal ($\sim$120\,ms for domain classification) rather than the sum.

\subsection{LoRA Memory Efficiency}

\Cref{tab:lora_memory} shows the memory advantage of serving classifiers via LoRA adapters versus independent fine-tuned models.

\begin{table}[h]
\centering
\caption{Model memory: independent fine-tuned models vs.\ LoRA adapters (ModernBERT base, 150M params)}
\label{tab:lora_memory}
\begin{tabular}{lrr}
\toprule
\textbf{Tasks ($n$)} & \textbf{Independent (MB)} & \textbf{LoRA (MB)} \\
\midrule
1 & 573   & 573 \\
3 & 1{,}719 & 574 \\
6 & 3{,}438 & 575 \\
\bottomrule
\end{tabular}
\end{table}

At $n = 6$, the LoRA architecture requires $\sim$6$\times$ less model memory (one base model + six $\sim$0.2\,MB adapters vs.\ six full model copies).
Each task still requires its own forward pass; latency reduction comes from \emph{parallel execution} of classifiers rather than from LoRA itself (\Cref{sec:lora_mom}).

\subsection{Decision Engine Overhead}

Decision evaluation adds negligible latency:
$< 0.1$\,ms for 10 decisions with 3 conditions each;
$< 0.5$\,ms for 100 decisions with 5 conditions each.
This confirms that the $O(M \cdot L_\text{max})$ complexity is dominated by signal extraction.

\subsection{Composable Orchestration Across Deployment Scenarios}

A key claim of this work is that the same architecture serves diverse deployment scenarios through configuration.
\Cref{tab:deployment_scenarios} demonstrates how different signal-decision-plugin compositions address different requirements:

\begin{table}[h]
\centering
\caption{Composable signal orchestration across deployment scenarios. Each scenario activates a different subset of the eleven signal types, selection algorithms, and plugin chains---using the same system binary and architecture.}
\label{tab:deployment_scenarios}
\begin{tabular}{p{2.5cm}p{3cm}p{2.5cm}p{3.5cm}}
\toprule
\textbf{Scenario} & \textbf{Active Signals} & \textbf{Selection} & \textbf{Key Plugins} \\
\midrule
Privacy-regulated (healthcare) & authz, domain, language & Static (compliant models only) & Strict PII redaction, no caching, audit logging \\
Cost-optimized (developer tool) & complexity, embedding, keyword & AutoMix cascade & Aggressive semantic cache, header mutation for LoRA adapter \\
Multi-cloud enterprise & domain, modality, authz & Latency-aware & Multi-endpoint failover, provider auth factory, system prompt injection \\
Multi-turn assistant & embedding, feedback, preference & Elo with session pin & Responses API state, memory retrieval, RAG injection \\
\bottomrule
\end{tabular}
\end{table}

\subsection{End-to-End Routing Correctness}

The end-to-end test framework validates routing behavior across eight scenario profiles:

\begin{table}[h]
\centering
\caption{End-to-end test profiles}
\label{tab:e2e_profiles}
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Profile} & \textbf{Validated Behavior} \\
\midrule
Multi-endpoint      & Multi-provider routing with weighted distribution and failover across heterogeneous backends \\
Multi-provider auth & Provider-specific auth injection (API key, OAuth2, cloud IAM) via authorization factory \\
AuthZ-RBAC          & Role-based model access (admin/premium/free tiers) with authz signal \\
ML model selection  & KNN, KMeans, SVM, MLP selection accuracy on held-out queries \\
Keyword routing     & Keyword signal matching with AND/OR/NOR combinators \\
Embedding routing   & Embedding similarity thresholds and confidence-based decision selection \\
RAG + Responses API & Context retrieval, injection, and stateful multi-turn via Responses API \\
Routing strategies  & Static, Elo, RouterDC, AutoMix, Hybrid algorithm comparison \\
\bottomrule
\end{tabular}
\end{table}

Each profile validates correct model selection, safety enforcement (jailbreak blocked, PII detected), cache behavior (hits after similar queries), multi-provider routing (correct endpoint resolution and auth injection), and header propagation.

\subsection{Semantic Cache Effectiveness}

At a similarity threshold $\theta = 0.92$:
exact-match queries achieve 100\% hit rate with $< 5$\,ms lookup latency;
paraphrased queries achieve 60--80\% hit rate depending on paraphrase distance.
Cache hits eliminate backend model invocation entirely, reducing per-request cost to embedding computation only.

\subsection{Open Evaluation}

Detailed model selection quality comparisons across algorithms, HaluGate detection accuracy on standard benchmarks (HaluEval, FActScore), and large-scale routing quality evaluation are under preparation in collaboration with the RouterArena team.
