\section{Related Architectures}
\label{sec:related}

\paragraph{Conditional computation inside a model.}
Adaptive mixtures and conditional-computation methods established the use of a
learned gate to allocate examples to specialized computation
\citep{jacobs_1991_mixture,bengio_2013_conditional}. Sparsely gated
Mixture-of-Experts (MoE), GShard, and Switch Transformers scale that principle
while activating only a subset of parameters for each input
\citep{shazeer_2017_moe,lepikhin_2020_gshard,fedus_2021_switch}. Their experts
typically share a checkpoint, training loop, tensor interface, and deployment
boundary. MoM retains conditional computation but moves the boundary outward:
a constituent can be an independently versioned local model, a private
specialist, or a closed remote service. MoM does not require joint
differentiability and often forgoes it; in return, it permits independent
evolution, heterogeneous placement, and explicit policy and failure boundaries.

\paragraph{Cascades and model routing.}
Cost-aware cascades and learned routers choose among models under quality and
resource objectives
\citep{aggarwal_2023_automix,chen_2023_frugalgpt,ding_2024_hybrid,
ong_2024_routellm}; RouterBench supplies a routing evaluation framework
\citep{hu_2024_routerbench}. Model-GLUE instead studies model selection and
parameter or output aggregation over a model zoo \citep{zhao_2024_modelglue}.
These methods provide important controllers and protocols. A general MoM also
admits parallel calls, verification, synthesis, and bounded workflows, and
places contracts and deployment identity inside the model lifecycle.

Closer whole-model precedents route among domain specialists, train asynchronous
language-model mixtures, form domain-specific consensus, or orchestrate models
and agents \citep{simonds_2024_modem,filippova_2025_smalltalk,
wang_2025_telemom,guo_2025_moma,pecerskis_2026_nsed}. Brick and NSED
independently use the Mixture-of-Models name; HyDRA and RouteBalance study
closely related capability routing and serving-aware scheduling
\citep{massa_2026_brick,garg_2026_hydra,da_2026_routebalance}. Fugu is a
particularly close contemporary system: trained orchestrator models dynamically devise
agentic scaffolds over a model team and expose the result through a model-like
interface \citep{tang_2026_fugu}. Our focus is the open infrastructure boundary
around such behavior---a finite execution algebra, portable identity,
environment binding and lock, attributable traces, and one lifecycle across
training, evaluation, and serving.

\paragraph{Ensembles and inference-time collaboration.}
Ensembling, response fusion, and Mixture-of-Agents exploit complementary outputs
from several models \citep{jiang_2023_llmblender,wang_2024_moa,li_2025_selfmoa}.
The cited methods evaluate a fixed composition topology. In our formulation,
parallel deliberation is one topology in a larger trace space, and a controller
may choose another under explicit resource and policy constraints. The trace
also makes its additional inference compute visible.

\paragraph{Compound AI systems and language-model programming.}
Compound AI systems frame application intelligence as the interaction of
models, retrieval, tools, and control logic \citep{kandogan_2024_compound}.
Language-model programming systems likewise provide constraints, control flow,
optimization, and efficient execution for multi-call programs
\citep{beurerkellner_2022_lmql,khattab_2023_dspy,zheng_2023_sglang}.
MoM narrows that broad systems view into a model-facing abstraction with a
bounded graph, explicit optimization variables, and portable identity. An agent
can invoke a MoM or request a bounded workflow, while the infrastructure layer
retains authority over admissibility and accounting. Programming systems can be
authoring frontends or execution backends; they do not by themselves define the
deployment identity proposed here.

\paragraph{Serving graphs, model packages, and runtimes.}
KServe InferenceGraph offers declarative sequence, switch, ensemble, and splitter
nodes behind one endpoint \citep{kserve_2026_inferencegraph}. Hugging Face model
repositories offer revisioned model coordinates; MLflow packages signatures and
loading flavors; and OCI manifests provide content-addressed distribution
\citep{huggingface_2026_models,mlflow_2026_model_format,
oci_2025_image_spec}. ONNX most clearly separates a typed model IR from
capability-negotiated execution providers
\citep{onnx_2026_ir,onnxruntime_2026_execution_providers}, while GGUF exemplifies
a self-describing materialized edge artifact \citep{gguf_2026_spec}. None alone
defines a composite model whose constituents may also be closed services. The
proposed lifecycle retains native leaf formats while adding a
preference-conditioned trace distribution, hard admissibility, separate
logical and physical identities, and trace-linked evidence. Serving runtimes
address complementary memory management, model multiplexing, scheduling and
migration, and multi-tenant adapter serving; examples include vLLM, MuxServe,
Llumnix, and Punica
\citep{kwon_2023_vllm,duan_2024_muxserve,sun_2024_llumnix,chen_2023_punica}.
A MoM artifact can bind to these systems without embedding mutable endpoints or
credentials.

\paragraph{Evolution of vLLM Semantic Router.}
Prior project papers developed signal-driven routing, workload--router--pool
optimization, a routing policy DSL, and agent orchestration
\citep{liu_2026_semanticrouter,chen_2026_wrp,liu_2026_conflictfree,chen_2026_orchestration}.
The earliest of these used ``MoM'' for mixture-of-modality deployments; here,
modality is one dimension of the broader catalog of independently versioned
models. Those mechanisms remain components of the reference engine. This paper
changes the unit of analysis from a routing decision to a versioned logical
model whose learned and authored state spans a pool, a control policy, an execution
topology, and contracts, together with an explicit deployment binding.

\section{Limitations and Research Agenda}
\label{sec:limitations}

The architecture covers finite, typed, resource-bounded graphs; it does not
make unbounded agency safe or guarantee a feasible binding. Mutable closed
services limit exact replay, the current system results do not establish the
matched-budget scaling hypothesis, and cross-provider or CPU--GPU--NPU portability
still requires controlled validation. Security, sovereignty, and grounding are
only as reliable as their evidence, enforcement, and checkers. Within those
boundaries, the research agenda rests on four claims with concrete validation
criteria.

\paragraph{Useful scale is complementarity at fixed active compute.}
A larger pool matters only when it adds capability, failure, resource, or
deployment diversity that a learned policy can exploit within the same call,
token, cost, energy, or latency budget. Matched-budget studies must therefore
report the strongest constituent,
oracle headroom, controller realization, repeated runs, error correlations, and
topology ablations as specified in Section~\ref{sec:required-experiments}.
Parallel panels and additional rounds count as active compute, not hidden model
capacity. The scaling curve of interest is attainable utility against catalog
diversity and active work; merely spending more inference is not conditional
scaling.

\paragraph{Logical model identity must survive physical change.}
If a MoM is to move between a robot, an edge site, a sovereign cluster, and a
managed service, its semantics cannot be identified with any one endpoint or
runtime. The bundle therefore implies a shared conformance suite:
canonicalization, signatures, operator capability negotiation, compatibility
tests, and independent engine implementations. A binder and resolver may
rebind logical references across CPU, GPU, NPU, and provider environments, but
every substitution must remain visible in the binding, lock, and evaluation. This is
especially important for mutable services whose weights, safety layers, prices,
limits, or aliases can change without exposing a checkpoint hash; a lock
records the best available evidence, while canaries and explicit revalidation
measure the remaining drift.

\paragraph{Lifecycle optimization must coordinate logical and physical state.}
Signals, control policies, pools, prompts, topologies, contracts, and placements
are optimized from different evidence and evolve on different timescales. Improving one in
isolation can destabilize another: a more aggressive gate changes fleet load,
a new binding changes latency and fallback behavior, and a new judge can change
the apparent reward. MoM lifecycle optimization therefore needs artifact- and
binding-level promotion, chronological evaluation, invariants, and rollback
\citep{chen_2026_wrp}. The research target is not convergence of one controller,
but stable improvement of realized utility and constraint satisfaction under
deployment drift.

\paragraph{Sovereignty and governance are properties of the trace.}
Sovereignty is not a binary choice between a local model and a public cloud.
Real deployments combine devices, jurisdiction-bound data centers, regional
providers, open weights, and selectively approved global services. Their
requirements span data and model control, residency, encryption-key control,
operations, supply chain, audit, and exit paths
\citep{european_commission_2026_sovereignty}. A MoM compiler should translate
such a profile into typed admissibility constraints over models, operators,
data paths, and bindings; cloud fallback must never silently override them.
The same principle applies to security, grounding, and preference governance.
Signals can guide execution but do not prove confidentiality or factuality,
and aggregate utility can hide systematically poor service for a group.
Least-privilege operators, secret separation, explicit boundary crossings,
per-group constraints, and worst-case evaluation make governance inspectable
on the realized trace rather than aspirational at the endpoint.

Progress on these four claims would turn model composition from a collection of
recipes into a model ecosystem with measurable scaling behavior, portable
identity, and a durable lifecycle.
